---
title: "DoubleML Tutorial"
author: "Zachary Goldstein"
date: "2023-05-12"
output:
  pdf_document:
     latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This is a tutorial for how to use the DoubleML package in R. The package implements Double/Debiased Machine Learning for estimating causal effects.

# Load Libraries
```{r,warning=F,message=F}
library(DoubleML)
library(tidyverse)
library(causaldata)
```

# Example Data

For this tutorial, we'll be using data from the Current Population Survey.

The data lets us study the effect of participation in a job-training program on
future wages.

The data is observational, it is not from an experiment and we can not assume
that program participation is independent of the potential outcomes.

Besides the main treatment and outcome variables, we have data on some covariates that we hypothesize to be related to participation
and wages, including pre-program wages, race, age, educational attainment, and 
marital status.

For the purposes of this tutorial, we are going to assume that the observed covariates are sufficient to control for confounding. (comparisons with
randomized experiment data have shown this may not actually be the case)

```{r}
df = cps_mixtape
df %>% head()
```

## Feature engineering

```{r}
df = df %>% 
  mutate(age_squared = age**2,
         age_under_35 = as.numeric(age<35),
         black_or_hispanic = as.numeric(black==1|hisp==1),
         # Average the income from the 2 years of data pre-program
         avg_pre_income = re74+re75/2,
         # Binary indicator for 12+ years of educational attainment
         educ_12_plus = as.numeric(educ>=12),
         )
```


# Partially Linear Model

## Theoretical Overview

Let's start with a relatively simple option for Double Machine Learning, the 
partially linear model. A partially linear model assumes a linear relationship
between the treatment and outcome, but makes no such assumptions about the 
relationships between the covariates and the outcome. We can use non-parametric
machine learning methods of our choice to model the relationship between the 
covariates and the outcome.

The model is as follows:
$$
Y = D\theta_0 + g_0(X) + \zeta
\\
D = m_0(X) + V
\\
E[\zeta|D,X] = 0
\\
E[V|X] = 0
$$

*Y* is the continuous outcome variable **re78**, real earnings in 1978.
*D* is the binary treatment variable **treat** corresponding to job training 
program participation.
$\theta_0$ is the effect of the treatment on the outcome.
$g_0$ is a possibly non-linear function of the covariates *X* that models their
relationship with the outcome.
*m* is the propensity score function that predictions the probability of 
program participation given the covariates *X*.
𝜁and *V* are random error terms.

## DoubleMLPLR function

### "New" method
The DoubleML R package implementation for the partially linear regression model is 
the function DoubleMLPLR. To create a new model, we use the "new" method,
as follows:

```{r}
plm = DoubleMLPLR$new(
  data = ,
  ml_l = ,
  ml_m = ,
  ml_g = ,
  n_folds = 5,
  n_rep = 1,
  score = "partialling out",
  dml_procedure = "dml2",
  draw_sample_splitting = TRUE,
  apply_cross_fitting = TRUE
)
```
#### Score

The score is which formula we use for estimating $\theta_0$, the effect of the treatment
on the outcome. The package gives us two options for the partially linear model score, the 
"partialling out" score and the "IV-type" score. In the former, we
set `score='partialling out'` and set learners `ml_l` and `ml_m`. In the latter,
we set `score='IV-type` and set learners `ml_l`, `ml_m`, and `ml_g`.

The partialling out approach uses the following formula:
$$
\check{\theta}_0 = \left(\frac{1}{n} \sum_{i\in I} \hat{V}_i \hat{V}_i \right)^{-1} \frac{1}{n} \sum_{i\in I} \hat{V}_i (Y_i - \hat{\ell}_0(X_i))
$$

And the IV-type uses the following formula:
$$\check{\theta}_0 = \left(\frac{1}{n} \sum_{i\in I} \hat{V}_i D_i\right)^{-1} \frac{1}{n} \sum_{i\in I} \hat{V}_i (Y_i - \hat{g}_0(X_i))
$$

TODO: EXPLAIN MORE ABOUT FORMULAS, ORTHOGONALIZATION, PROS/CONS OF EACH

#### Learners

#### DML Procedure

#### Sample Splitting

discuss n_rep

#### Cross Fitting

discuss n_folds



# Scratch

DoubleMLPLR$new()

DoubleMLPLR$set_ml_nuisance_params()

DoubleMLPLR$tune()

DoubleMLPLR$clone()

Inherited methods
DoubleML::DoubleML$bootstrap()
DoubleML::DoubleML$confint()
DoubleML::DoubleML$fit()
DoubleML::DoubleML$get_params()
DoubleML::DoubleML$learner_names()
DoubleML::DoubleML$p_adjust()
DoubleML::DoubleML$params_names()
DoubleML::DoubleML$print()
DoubleML::DoubleML$set_sample_splitting()
DoubleML::DoubleML$split_samples()
DoubleML::DoubleML$summary()

If I want even more content, I can discuss IV or cluster data or interactive regression.

# Citations

Bach, P., Chernozhukov, V., Kurz, M. S., and Spindler, M. (2021), DoubleML - An Object-Oriented Implementation of Double Machine Learning in R, arXiv:2103.09603.

Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C., Newey, W., & Robins, J. (2018). Double/debiased machine learning for treatment and structural parameters.

